A Classical Density-Functional Theory for Describing Water Interfaces

### 2012-07-31 papers/water-SAFT/paper.tex
updated ab inition MD citation Tested everything on node10.

A large fraction of interesting chemistry---including all of molecular
biology---takes place in aqueous solution.  However, while quantum
chemistry now enables us to calculate the ground state energies of
large molecules in vacuum, prediction of the free energy of even the
smallest molecules in the presence of a solvent poses a continuing
challenge due to the complex structure of a liquid and the
computational cost of \emph{ab initio} molecular
dynamics~\cite{car1985, grossman2004}.  The current state-of-the art
in \emph{ab initio} molecular dynamics is limited to around 64 water
molecules per unit cell, which places the minimum solute concentration
at 0.84 molar and allows contact between the second solvation shell of
neighboring solutes~\cite{izvekov2005, choe2007}\fixme{Look up more
  recent numbers on this!}.  On top of this, standard \emph{ab initio}
methods using classical molecular dynamics without van der Waals
corrections strongly over-structure water, to the point that ice melts
at over 400K\cite{yoo2009phase}!  There has been a recent flurry of
publications implicating van der Waals effects as significant in
reducing this over-structuring\cite{lin2009importance,
  wang2011density, mogelhoj2011ab, jonchiere2011van}.  It has also
been found that the inclusion of nuclear quantum effects can provide
similar improvements \cite{morrone2008nuclear}.  Each of these
corrections imposes an additional computational burden on an approach
that is already feasible for only a very small number of water
molecules. A more efficient approach is needed in order to study
nanoscale and hydrophobic solutes.

@@

We begin by studying a single hydrophobic rod immersed in water.
% We
% model the density on a grid with spacing of about 0.01~nm, which gives
% the free energy of hydration to a precision of better than a percent,
% and the density at contact to a few percent precision.
We show in Figure~\ref{fig:energy-vs-diameter} the excess chemical
potential at room temperature, scaled by the solvent-accessible
surface area of the hard rod, plotted as a function of hard-rod
radius, which we define as the radius from which water is excluded.
For rods with radius larger than 0.5~nm or so, this reaches a value of
75~mN/m, which is slightly higher than the bulk surface, which will be
approached from above in the limit of very large rods \fixme{this is a
terrible sentence, want to say something more concrete about free
energy / area dropping (getting data) DJR}.  For very small
radius (less than about 0.1~nm) the relationship between the free
energy per area and radius is linear, with the slope required by the
contact-value theorem (see Equation~\ref{contactvaluethm}).

>>>>

A large fraction of interesting chemistry---including all of molecular
biology---takes place in aqueous solution.  However, while quantum
chemistry now enables us to calculate the ground state energies of
large molecules in vacuum, prediction of the free energy of even the
smallest molecules in the presence of a solvent poses a continuing
challenge due to the complex structure of a liquid and the
computational cost of \emph{ab initio} molecular
dynamics~\cite{car1985, grossman2004}.  The current state-of-the art
in \emph{ab initio} molecular dynamics is limited to a few hundred water
molecules per unit cell~\cite{lewis2011doesnitric}.  On top of this,
standard \emph{ab initio}
methods using classical molecular dynamics without van der Waals
corrections strongly over-structure water, to the point that ice melts
at over 400K\cite{yoo2009phase}!  There has been a recent flurry of
publications implicating van der Waals effects as significant in
reducing this over-structuring\cite{lin2009importance,
  wang2011density, mogelhoj2011ab, jonchiere2011van}.  It has also
been found that the inclusion of nuclear quantum effects can provide
similar improvements \cite{morrone2008nuclear}.  Each of these
corrections imposes an additional computational burden on an approach
that is already feasible for only a very small number of water
molecules. A more efficient approach is needed in order to study
nanoscale and hydrophobic solutes.

@@

We begin by studying a single hydrophobic rod immersed in water.
% We
% model the density on a grid with spacing of about 0.01~nm, which gives
% the free energy of hydration to a precision of better than a percent,
% and the density at contact to a few percent precision.
We show in Figure~\ref{fig:energy-vs-diameter} the excess chemical
potential at room temperature, scaled by the solvent-accessible
surface area of the hard rod, plotted as a function of hard-rod
radius, which we define as the radius from which water is excluded.
For rods with radius larger than 0.5~nm or so, this reaches a maximum value of
75~mN/m, which is slightly higher than the bulk surface.  In the limit
of very large rods, this value will decrease and approach the bulk
surface value.  For very small radius (less than about 0.1~nm) the
relationship between the free energy per area and radius is linear,
with the slope required by the contact-value theorem (see Equation~\ref{contactvaluethm}).
